---
title: paste("Individual Report for Patient", directory)
author: Anish Shah, MD
output: pdf_document
params:
  data: "sample"
editor_options: 
  chunk_output_type: console
---

```{r setup and script args, global_options, include=FALSE}
# Start time
alpha_time <- Sys.time()

# Knitr options
knitr::opts_chunk$set(
  cache = TRUE,
  eval = TRUE,
  include = FALSE,
  options("scipen" = 999, "digits = 2")
)


## Script arguments

# Name of folder from command line
# Will use this for file.path function
args <- commandArgs(TRUE)
name <- args[1]
name <- "UI7616" # Temporary

## Source Libraries
source("1_libraries.R")


## Source Files

# Folder locations
folder <- file.path(getwd(), "proc_patients", name)
raw_folder <- file.path(getwd(), "raw_patients")

# Read in the files
source("2_intake.R") 

## Source Tidying
source("3_tidy.R")

# End of process
omega_time <- Sys.time()
cat("Time to load libraries, intake data, and tidy it...")
print(omega_time - alpha_time)
```

## Overview of HRV data

The processed HRV from MATLAB toolbox has created a CSV file with all HRV variables computed. Below is a description:

- Length of recording: `r nrow(rawfile) * increment / 60 / 60` hours
- Increment window: `r as.numeric(hrvparams[hrvparams$Tab1=="increment",2])`
- Window length: `r as.numeric(hrvparams[hrvparams$Tab1 == "windowlength", 2])`
- Data not analyzed: `r as.numeric(removed$PercentNotAnalyzed[1])`%

The data columns are: `r names(df)`

## First Pass Visualization

```{r HRV graphs}
# Visualize data
all <-
  gather(df, key = "hrv", value = "value", -time) %>%
  ggplot(., aes(x = time, y = value, color = hrv)) +
  geom_smooth() +
  scale_x_datetime(date_breaks = "3 hours", date_labels = "%H") +
  geom_vline(xintercept = cath_time, linetype = 4) +
  facet_wrap(~hrv, scales = "free_y") +
  scale_color_viridis(discrete = TRUE) + 
  theme_minimal() +
  theme(legend.position = "none") +
  labs(
    title = "Heart Rate Variability",
    x = "Time in Hours",
    y = "HRV"
  )

psa <- 
  gather(df[c("time","HF","LF","VLF","TP","LFHF")], key = "hrv", value = "value", -time) %>%
  ggplot(., aes(x = time, y = value, color = hrv)) +
  scale_x_datetime(date_breaks = "1 hour", date_labels = "%H") +
  geom_vline(xintercept = cath_time, linetype = 4) +
  geom_smooth() +
  scale_color_viridis(discrete = TRUE) + 
  theme_minimal() +
  labs(
    title = "Power spectral analysis",
    x = "Time in Hours",
    y = "HRV"
  )

accel <- 
  gather(df[c("time","AC", "DC")], key = "hrv", value = "value", -time) %>%
  ggplot(., aes(x = time, y = value, color = hrv)) +
  scale_x_datetime(date_breaks = "1 hour", date_labels = "%H") +
  geom_vline(xintercept = cath_time, linetype = 4) +
  geom_smooth() +
  scale_color_viridis_d() + 
  theme_minimal() +
  labs(
    title = "Acceleration and Deceleration Capacity",
    x = "Time in Hours",
    y = "log(HRV)"
  )

timeDomain <-
  gather(df[c("time","SDNN","RMSSD","PNN50")], key = "hrv", value = "value", -time) %>%
  ggplot(., aes(x = time, y = value, color = hrv)) +
  scale_x_datetime(date_breaks = "1 hour", date_labels = "%H") +
  geom_vline(xintercept = cath_time, linetype = 4) +
  geom_smooth() +
  scale_color_viridis_d() + 
  theme_minimal() +
  labs(
    title = "Time Domain",
    x = "Time in Hours",
    y = "HRV"
  )

entropy <-
  gather(df[c("time","SampEn", "ApEn")], key = "hrv", value = "value", -time) %>%
  ggplot(., aes(x = time, y = value, color = hrv)) +
  scale_x_datetime(date_breaks = "1 hour", date_labels = "%H") +
  geom_vline(xintercept = cath_time, linetype = 4) +
  geom_smooth() +
  scale_color_viridis_d() + 
  theme_minimal() +
  labs(
    title = "Time Domain",
    x = "Time in Hours",
    y = "HRV"
  )
  
# Save the figures
ggsave(file.path(directory, "all_hrv.png"), plot = all)
ggsave(file.path(directory, "timeDomain_hrv.png"), plot = timeDomain)
ggsave(file.path(directory, "entropy_hrv.png"), plot = entropy)
ggsave(file.path(directory, "psa_hrv.png"), plot = psa)
ggsave(file.path(directory, "accel_hrv.png"), plot = accel)
```