---
title: "Depression, HRV, and CAD Pilot Study"
subtitle: "the Emory Cardiovascular Biobank" 
author: 
- Anish Shah, MD^[Department of Medicine, School of Medicine, Emory University]
- Amit Shah, MD, MSCR^[Divison of Cardiology, Department of Medicine, Emory University]
- Alvaro Alonso, MD, PhD^[Department of Epidemiology, Rollins School of Public Health, Emory University]
fontsize: 10pt
output: 
  beamer_presentation:
    colortheme: "beaver"
    slide_level: 3
  latex_engine: xelatex
header-includes:
  - \usepackage{dcolumn}
  - \usepackage{float}
  - \usepackage{graphicx} 
  - \usepackage{booktabs}
  - \usepackage{longtable}
  - \usepackage{array}
  - \usepackage{multirow}
  - \usepackage{wrapfig}
  - \usepackage{colortbl}
  - \usepackage{pdflscape}
  - \usepackage{tabu}
  - \usepackage{threeparttable}
always_allow_html: yes

---

```{r setup and script args, global_options, include=FALSE}
# Start time
alpha_time <- Sys.time()

# Knitr options
knitr::opts_chunk$set(
  cache = TRUE,
  warning = FALSE,
  eval = TRUE,
  echo = FALSE,
  include = TRUE,
  message = FALSE,
  dpi = 600,
  dev = "png",
  options("scipen" = 999, "digits" = 2),
  tinytex.verbose = TRUE,
  tidy = FALSE
)

options(xtable.comment = FALSE)
```

```{r load up data, include = FALSE}
# Libraries
source("1_libraries.R")

# Read in the files
source("2a_vivalnk.R")
source("2b_hrv.R")
source("2c_biobank.R")

# Source Tidying
source("3_tidy.R")

# End of process
omega_time <- Sys.time()
cat("Time to load libraries, intake data, and tidy it...")
print(omega_time - alpha_time)
```

# Overview

### Disclosures and Funding

- Emory University
- Emory Clinical Cardiovascular Research Institute
- Emory Program in Cardivascular Outcomes Research and Epidemiology
- Georgia Clinical & Translational Science Alliance
- NIH/NCATS UL1TR002378, TL1 TL1TR002382

### Emory Cardiovascular Biobank

- Ongoing prospective cohort of patients undergoing cardiac catherization
- Includes clinical history and biomarkers
- Includes psychological questionnaires (including depression by PHQ9)
- Over 7000 patients enrolled thus far

### Background

This is a pilot study examining the relationship between depression and CAD. We have shown using the non-linear HRV metric, Dyx, is a powerful predictor of CAD, and can also be a useful marker for Depression.


### Aims

1. Quantify how depression affects ANS function
2. Examine how HRV can predict obstructive (versus microvascular) CAD

# Descriptive analysis

## Population

### Demographic description

```{r demographics}
df <- demo
df$adm_reason <- df_txt$adm_reason
df$gend <- df_txt$gend
df$race <- df_txt$race
df$setting <- df_txt$setting


# Table 1
compareGroups(setting ~ gend + race + blbmi + adm_reason, data = df) %>%
  createTable(., hide = c(gend = "Female"), show.p.overall = FALSE) %>%
  export2md(., format = "markdown", caption = "Patient description")
```

## Depression

### Depression scoring

- Each patient is given a questionnaire, the PHQ9
- The scores are validated and suggest severity/category of depression
- Scores >= 10 are considered moderate to severe depression, and accepted cut-off

### Depression table

```{r depression scoring}
df <- inner_join(demo, psych, by = "patid")
df$adm_reason <- df_txt$adm_reason
df$gend <- df_txt$gend
df$race <- df_txt$race
df$sad %<>% factor()
levels(df$sad) <- c("Low Depression", "High Depression")

# Depression comparison
compareGroups(data = df, sad ~ gend + race + adm_reason) %>%
  createTable(., hide = c(gend = "Female"), show.p.overall = FALSE) %>%
  export2md(., format = "markdown", caption = "Depression scoring")
```

## Coronary artery disease

### Cardiac catherization

- Every patient presents for cardiac catherization to be enrolled
- Are either inpatient or outpatient
- Etiology: pre-op, heart transplant, UA, NSTEMI, STEMI, positive stress test
- scored by angiographic severity indices - CASS and Gensini scores

### CAD Severity

```{r CAD overview}
df <- inner_join(demo, angio_scores, by = "patid")
df$adm_reason <- df_txt$adm_reason
df$gend <- df_txt$gend
df$race <- df_txt$race

# CAD scoring systems
compareGroups(data = df, ~ cass50 + cass70 + gensini + stenosis) %>%
  createTable(., show.p.overall = FALSE) %>%
  export2md(., format = "markdown", caption = "CAD Severity Scores")
```

## Heart rate variability

### Overview of HRV

- ECG data was collected using the VivaLNK patch
- this records data for up to 72 hours
- ECG was started the AM of LHC, and continued for several hours after event
- HRV is generated through the Emory HRV Toolbox 
- Frequency domain was log-transformed
- HRV was blocked into averaged 1-hour segments for analysis

### Quality of HRV data

```{r}
df <- hrv_quality

# HRV quality
compareGroups(data = df, ~ Duration + PercentNotAnalyzed + PercentLowQualityWind) %>%
  createTable(., show.p.overall = FALSE) %>%
  export2md(., format = "markdown", caption = "HRV quality")
```

### Overview of all HRV measures {.allowframebreaks}

```{r, descriptive HRV}
df <- hrv_proc

# Time domain
compareGroups(data = df, ~ BPM + SDNN + RMSSD + PNN50) %>%
  createTable(., show.p.overall = FALSE) %>%
  export2md(., format = "markdown", caption = "Time Domain")

# Frequency domain
compareGroups(data = df, ~ HF + LF + VLF + ULF + TP + LFHF) %>%
  createTable(., show.p.overall = FALSE) %>%
  export2md(., format = "markdown", caption = "Frequency Domain")

# Other measures
compareGroups(data = df, ~ AC + DC + SampEn + ApEn) %>%
  createTable(., show.p.overall = FALSE) %>%
  export2md(., format = "markdown", caption = "Additional Measures")
```

```{r HRV graphs}
df <- hrv_long

# Visualize
ggplot(df, aes(x = clock, y = values, color = hrv)) +
  geom_smooth(se = TRUE) +
  geom_ribbon(aes(ymin = ymin, ymax = ymax)) +
  scale_x_datetime(date_breaks = "4 hours", date_labels = "%H") +
  facet_wrap(~hrv, scales = "free_y") +
  scale_color_viridis_d() +
  theme_minimal() +
  theme(
    legend.position = "none"
  ) +
  labs(
    title = "HRV",
    x = "Time (hour of day)",
    y = "HRV"
  )

df$clock <- hour(df$clock) + minute(df$clock) / 60
df <- na.omit(df)

# Visualizing all the data will be expensive
# Even reducing to just a few patients may be exhaustingly slow
gTimeHRV <-
  ggplot(subset(df, patid %in% c("06931", "06955", "07027", "07052", "06944")), aes(x = clock, y = values)) +
  facet_wrap(~ hrv, ncol = 2) +
  stat_summary(aes(group = hrv, y = values), fun.data = "mean_cl_normal", geom = "ribbon", alpha = 1 / 5) +
  stat_summary(aes(group = hrv, colour = hrv, y = values), fun.y = mean, geom = "line") +
  labs(
    title = "Heart Rate Variability over 24-hours",
    x = "Time of Day (hours)",
    y = "Mean and standard error of the mean",
    colour = "HRV measures"
  ) +
  scale_colour_viridis_d() + 
  theme_minimal() + 
  theme(
    legend.position = "none"
  ) 
```


# Aim 1: Relationship between Depression and ANS Dysfunction

```{r exploring depression and HRV}
df <- inner_join(psych[c("patid", "sad")], hrv_proc, by = "patid")

# Distribution of sadness
ks.test(df$HF[df$sad == 1], df$HF[df$sad == 0])
ks.test(df$LF[df$sad == 1], df$LF[df$sad == 0])
ks.test(df$VLF[df$sad == 1], df$VLF[df$sad == 0])
ks.test(df$ULF[df$sad == 1], df$ULF[df$sad == 0])
ks.test(df$TP[df$sad == 1], df$TP[df$sad == 0])
ks.test(df$LFHF[df$sad == 1], df$LFHF[df$sad == 0])
ks.test(df$AC[df$sad == 1], df$AC[df$sad == 0])
ks.test(df$DC[df$sad == 1], df$DC[df$sad == 0])
ks.test(df$LF[df$sad == 1], df$LF[df$sad == 0])
ks.test(df$SampEn[df$sad == 1], df$SampEn[df$sad == 0])
ks.test(df$ApEn[df$sad == 1], df$ApEn[df$sad == 0])
ks.test(df$BPM[df$sad == 1], df$BPM[df$sad == 0])
```



```{r circadian depression and T-test}
# Df for visualization
df <-
  hrv_blocks %>%
  pivot_longer(., names_to = "hrv", values_to = "value", -c(patid, hour)) %>%
  inner_join(psych[c("patid", "sad")], ., by = "patid") %>%
  na.omit() %>%
  group_by(sad, hour, hrv) %>%
  dplyr::summarise(mean = mean(value), n = length(value), sd = sd(value), se = sd(value) / sqrt(length(value)))

# Have P values for significant differences handy (not currently working)
overall_t <-
  inner_join(subset(psych, !is.na(sad)), hrv_blocks, by = "patid") %>%
  na.omit() %>%
  do(
    NN = t.test(NN ~ sad, data = .)$p.value,
    SDNN = t.test(SDNN ~ sad, data = .)$p.value,
    RMSSD = t.test(RMSSD ~ sad, data = .)$p.value,
    PNN50 = t.test(PNN50 ~ sad, data = .)$p.value,
    ULF = t.test(ULF ~ sad, data = .)$p.value,
    VLF = t.test(VLF ~ sad, data = .)$p.value,
    LF = t.test(LF ~ sad, data = .)$p.value,
    HF = t.test(HF ~ sad, data = .)$p.value,
    LFHF = t.test(LFHF ~ sad, data = .)$p.value,
    TP = t.test(TP ~ sad, data = .)$p.value,
    AC = t.test(AC ~ sad, data = .)$p.value,
    DC = t.test(DC ~ sad, data = .)$p.value,
    SampEn = t.test(SampEn ~ sad, data = .)$p.value,
    ApEn = t.test(ApEn ~ sad, data = .)$p.value,
  )
  gather(variable, value, -hour)
hourly_t$value <- as.numeric(hourly_t$value)

# Visualization of data faceted by HRV and grouped by ischemia
gCircadianSadness <-
  ggplot(df, aes(x = factor(hour), colour = sad)) +
  facet_wrap(~ hrv, scales = "free_y") +
  geom_point(aes(y = mean), size = 2, position = position_dodge(.3)) +
  scale_shape_manual(values = c(1, 16)) +
  geom_errorbar(aes(ymin = mean - se, ymax = mean + se), width = 0.2, position = position_dodge(.3))
  theme(panel.border = element_rect(colour = "black", fill = NA)) +
  labs(
    title = "Hourly HRV in those with depression",
    x = "Time in Hours",
    y = "Normalized HRV [SEM]",
    colour = "Depression",
  ) +
  scale_color_viridis(option = "cividis", discrete = TRUE, begin = .1, end = .9) +
  theme_minimal() +
  theme(
    plot.caption = element_text(hjust = 0),
    legend.position = "bottom",
    legend.box = "horizontal"
  )

# Plot this
gDiurnal

```

### Visualizing Differences in HRV by Depresion
```{r HRV x depression plot}
df <- 
  inner_join(hrv_proc, psych[c("patid", "sad")], by = "patid") %>%
  pivot_longer(., names_to = "hrv", values_to = "values", values_drop_na = TRUE, -c(patid, index, clock, hour, sad)) %>%
  na.omit()

# Visualize
ggplot(df, aes(x = clock, y = values, group = factor(sad), color = factor(sad))) + 
  geom_smooth(na.rm = TRUE) +
  scale_x_datetime(date_breaks = "4 hours", date_labels = "%H") +
  facet_wrap(~hrv, scales = "free_y") +
  scale_color_viridis_d(option = "E") +
  theme_minimal() +
  theme(legend.position = "none") + 
  labs(
    title = "HRV by Depression (sad = yellow)",
    x = "Time (hour of day)",
    y = "HRV"
  )
```

# Aim 2: Relationship between CAD and ANS Dysfunction

### Differences in HRV distribution by CAD

```{r}
df <- inner_join(angio_scores[c("patid", "stenosis")], hrv_proc, by = "patid")

ks.test(df$HF[df$stenosis == 1], df$HF[df$stenosis == 0])
ks.test(df$LF[df$stenosis == 1], df$LF[df$stenosis == 0])
ks.test(df$VLF[df$stenosis == 1], df$VLF[df$stenosis == 0])
ks.test(df$ULF[df$stenosis == 1], df$ULF[df$stenosis == 0])
ks.test(df$TP[df$stenosis == 1], df$TP[df$stenosis == 0])
ks.test(df$LFHF[df$stenosis == 1], df$LFHF[df$stenosis == 0])
ks.test(df$AC[df$stenosis == 1], df$AC[df$stenosis == 0])
ks.test(df$DC[df$stenosis == 1], df$DC[df$stenosis == 0])
ks.test(df$LF[df$stenosis == 1], df$LF[df$stenosis == 0])
ks.test(df$SampEn[df$stenosis == 1], df$SampEn[df$stenosis == 0])
ks.test(df$ApEn[df$stenosis == 1], df$ApEn[df$stenosis == 0])
ks.test(df$BPM[df$stenosis == 1], df$BPM[df$stenosis == 0])
```


### Visualizing Differences in HRV by CAD
```{r HRV x CAD plot}

df <- 
  inner_join(hrv_proc, angio_scores[c("patid", "stenosis")], by = "patid") %>%
  pivot_longer(., names_to = "hrv", values_to = "values", values_drop_na = TRUE, -c(patid, index, clock, hour, stenosis)) %>% 
  na.omit() %>%
  subset(., hrv %in% c("AC", "DC", "BPM", "HF", "LF", "TP", "LFHF", "PNN50", "VLF"))

# Visualize
ggplot(df, aes(x = clock, y = values, group = factor(stenosis), color = factor(stenosis))) + 
  geom_smooth(na.rm = TRUE) +
  scale_x_datetime(date_breaks = "4 hours", date_labels = "%H") +
  facet_wrap(~hrv, scales = "free_y") +
  scale_color_viridis_d(option = "E") +
  theme_minimal() +
  theme(legend.position = "none") + 
  labs(
    title = "HRV by Stenosis (+ = yellow)",
    x = "Time (hour of day)",
    y = "HRV"
  )
```

```{r, circadian CAD}
# Df for visualization
df <-
  hrv_blocks %>%
  pivot_longer(., names_to = "hrv", values_to = "value", -c(patid, hour)) %>%
  inner_join(angio_scores[c("patid", "stenosis")], ., by = "patid") %>%
  na.omit() %>%
  group_by(stenosis, hour, hrv) %>%
  dplyr::summarise(mean = mean(value), n = length(value), sd = sd(value), se = sd(value) / sqrt(length(value)))

# Have P values for significant differences handy (not currently working)
hourly_t <-
  inner_join(subset(angio_scores, !is.na(stenosis)), hrv_blocks, by = "patid") %>%
  group_by(hour) %>%
  na.omit() %>%
  do(
    NN = t.test(NN ~ stenosis, data = .)$p.value,
    SDNN = t.test(SDNN ~ stenosis, data = .)$p.value,
    RMSSD = t.test(RMSSD ~ stenosis, data = .)$p.value,
    PNN50 = t.test(PNN50 ~ stenosis, data = .)$p.value,
    ULF = t.test(ULF ~ stenosis, data = .)$p.value,
    VLF = t.test(VLF ~ stenosis, data = .)$p.value,
    LF = t.test(LF ~ stenosis, data = .)$p.value,
    HF = t.test(HF ~ stenosis, data = .)$p.value,
    LFHF = t.test(LFHF ~ stenosis, data = .)$p.value,
    TP = t.test(TP ~ stenosis, data = .)$p.value,
    AC = t.test(AC ~ stenosis, data = .)$p.value,
    DC = t.test(DC ~ stenosis, data = .)$p.value,
    SampEn = t.test(SampEn ~ stenosis, data = .)$p.value,
    ApEn = t.test(ApEn ~ stenosis, data = .)$p.value,
  )

# Visualization of data faceted by HRV and grouped by ischemia
  ggplot(df, aes(x = factor(hour), colour = stenosis)) +
  facet_wrap(~ hrv, scales = "free_y") +
  geom_point(aes(y = mean), size = 2, position = position_dodge(.3)) +
  scale_shape_manual(values = c(1, 16)) +
  geom_errorbar(aes(ymin = mean - se, ymax = mean + se), width = 0.2, position = position_dodge(.3)) +
  theme(panel.border = element_rect(colour = "black", fill = NA)) +
  labs(
    title = "Hourly HRV in those with CAD",
    x = "Time in Hours",
    y = "Normalized HRV [SEM]",
    colour = "CAD"
  ) +
  scale_color_viridis(option = "cividis", discrete = TRUE, begin = .1, end = .9) +
  theme_minimal()
```