---
title: "Depression and CAD"
author: "Anish Shah"
subtitle: Biobank analysis with sex and age
output: 
  pdf_document:
    latex_engine: xelatex
---

```{r setup and script args, global_options, include=FALSE}
# Knitr options
knitr::opts_chunk$set(
  cache = TRUE,
  eval = TRUE,
  include = TRUE,
  echo = FALSE,
  message = FALSE,
  warning = FALSE,
  options("scipen" = 999, "digits = 2")
)

# Source Libraries
source("1_libraries.R")
```

# Intake

```{r demographics}
# Read in demographic data
df_num <- read_csv("sex_sad_cad.csv", col_names = TRUE, guess_max = 30000)
#df_txt <- read_csv("sex_sad_cad_labels.csv", col_names = TRUE)

# Save the labels to match later
numVar <- names(df_num)
#txtVar <- names(df_txt)
```

```{r cath data}
# Cath data 
svar <- c(
"uniqueid",
"ang1dom",
"ang1edp",
"ang1ef",
"ang1cabg1",
"ang1cabg2",
"ang1cabg3",
"ang1cabg4",
"ang1cabg5",
"ang1cabg6",
"ang1cabg7",
"ang1cabg8",
"ang1cabg9",
"ang1cabg10",
"ang1cabg11",
"ang1cabg12",
"ang1cabg13",
"ang1cabg14",
"ang1results",
"ang1graf",
"ang1isr",
"ang1cabgsten1",
"ang1cabgsten2",
"ang1cabgsten3",
"ang1cabgsten4",
"ang1cabgsten5",
"ang1cabgsten6",
"ang1cabgsten7",
"ang1cabgsten8",
"ang1cabgsten9",
"ang1cabgsten10",
"ang1cabgsten11",
"ang1cabgsten12",
"ang1cabgsten13",
"ang1cabgsten14",
"ang1sten1",
"ang1sten2",
"ang1sten3",
"ang1sten4",
"ang1sten5",
"ang1sten6",
"ang1sten7",
"ang1sten8",
"ang1sten9",
"ang1sten10",
"ang1sten11",
"ang1sten12",
"ang1sten13",
"ang1sten14",
"ang1sten15",
"ang1sten16",
"ang1sten17",
"ang1sten18",
"ang1sten19",
"ang1sten20",
"ang1sten21",
"ang1sten22",
"ang1int1",
"ang1int2",
"ang1int3",
"ang1int4",
"ang1int5",
"ang1int6",
"ang1int7",
"ang1int8",
"ang1int9",
"ang1int10",
"ang1int11",
"ang1int12",
"ang1int13",
"ang1int14",
"ang1int15",
"ang1int16",
"ang1int17",
"ang1int18",
"ang1int19",
"ang1int20",
"ang1int21",
"ang1int22",
"ang1graftint1",
"ang1graftint2",
"ang1graftint3",
"ang1graftint4",
"ang1graftint5",
"ang1graftint6",
"ang1graftint7",
"ang1graftint8",
"ang1graftint9",
"ang1graftint10",
"ang1graftint11",
"ang1graftint12",
"ang1graftint13",
"ang1graftint14",
"ang1outcomes"
)

cath <- df_num[df_num$redcap_event_name == "enrollment_1_arm_1", svar]

#}}}
```

The Biobank data was extracted in full. Demographic and angiographic data was pulled from the initial enrollment. Depression scores were also extracted from initial enrollment.

# Tidy data

## Sex and age categories

The ages and sex of the patients were documented as well. We chose to group patients by age, based on <=60 and or >60 in age.

```{r, results = "asis"}
# Data frame for sex/age
df <- subset(df_num, redcap_event_name == "enrollment_1_arm_1")

# Breakdown is women/men at 3 age groups <= 55, 56 - 64, >= 65
df$ageBin <- ifelse(df$age <= 55, 0, ifelse(df$age < 65, 1, 2)) 

df$gend %<>% factor()
levels(df$gend) <- c("Women", "Men")

df$ageBin %<>% factor()
levels(df$ageBin) <- c("Young", "Middle", "Old")

# Save demo
df_demo <- df[c("uniqueid", "age", "gend", "ageBin")]

# Present
df_demo[,c(2:4)] %>% as.data.frame() %>% stargazer(., header = FALSE)
```

## Depression scores

Depression was scored by PHQ9. The overview is below.

```{r, results = "asis"}
# Data frame for depression
df <- subset(df_num, redcap_event_name == "enrollment_1_arm_1")

# Depression
phq9 <- c(
  "mdplea",
  "mddep",
  "mdsleep",
  "mdtired",
  "mdappt",
  "mdbad", 
  "mdconc",
  "mddead",
  "mdspeak"
)

# New data frame
df <- df[c("uniqueid", phq9)]

# SUm the PHq9
df %<>%
 mutate(phq = select(., mdplea:mdspeak) %>% rowSums(na.rm = TRUE)) 
df$phq[df$phq == 0] <- NA
df$phq[!is.na(df$phq)] <- df$phq[!is.na(df$phq)] - 9 # Since start at 0
df$phq[df$phq <= 0] <- 0

# CUt off of 10 for hte PHQ9
df$sad <- ifelse(df$phq >9, 1, 0)

df$sad %<>% factor()
levels(df$sad) <- c("Undepressed", "Depressed")

# Safe the table
df_sad <- df[c("uniqueid", "phq", "sad")]

df_sad[,c(2:3)] %>% as.data.frame() %>% stargazer(., header = FALSE)
```


## Coronary artery disease

All angiogram data was extracted from the biobank. Each patient was analyzed for first enrollment aniography data.

```{r}
# Data frame
df <- cath

# Stenoses are present?
df$stenosis <- df$ang1results

## CASS score generation

# CASS score? Generated as follows
  # 3 Major epicardioal vessles c- >70% stenosis = 1 point
    # LAD, LCX, RCA .... LM = LAD + LCX
  # Stenosis >50% in left main = 2 vessel disease = 2 point
    # Final score is sum of all points, analogous with 1-3 vessel dz
  # Can be defined as >70% stenosis or >50% stenosis

# ang1sten tell us which arteries have stenosis
# ang1sten(num) tell us the percent stenosis in that vessel
# ang1cabgsten(num) tell us percent stenosis in the graft
# ang1sten1 = left main
# ang1sten[2-5] = LAD prox_before, prox_after, mid, distal
# ang1sten[6-8] = LCX prox, mid, distal
# ang1sten[9-12] = RCA prox, mid, distal, distal_post

# THe epicardial vessels should be scores
df$lm <- as.numeric(df$ang1sten1)
df$lm[is.na(df$lm)] <- 0
df$lad <- as.numeric(apply(X = df[c("ang1sten2", "ang1sten3", "ang1sten4", "ang1sten5")], MARGIN = 1, FUN = max, na.rm = TRUE))
df$lad[df$lad == -Inf] <- 0
df$lad[is.na(df$lad)] <- 0
df$lcx <- as.numeric(apply(X = df[c("ang1sten6", "ang1sten7", "ang1sten8")], MARGIN = 1, FUN = max, na.rm = TRUE))
df$lcx[df$lcx == -Inf] <- 0
df$lcx[is.na(df$lcx)] <- 0
df$lcx %<>% as.numeric()
df$rca <- as.numeric(apply(X = df[c("ang1sten9", "ang1sten10", "ang1sten11", "ang1sten12")], MARGIN = 1, FUN = max, na.rm = TRUE))
df$rca[df$rca == -Inf] <- 0
df$rca[is.na(df$rca)] <- 0

# CASS-50 score is needed
df$cass50 <- 0
df$cass50[df$lad >= 50] <- df$cass50[df$lad >= 50] + 1
df$cass50[df$lcx >= 50] <- df$cass50[df$lcx >= 50] + 1
df$cass50[df$lm >= 50] <- 2
df$cass50[df$rca >= 50] <- df$cass50[df$rca >= 50] + 1
df$cass50[is.na(df$ang1results)] <- NA

# CASS-70 score is needed
df$cass70 <- 0
df$cass70[df$lad >= 70] <- df$cass70[df$lad >= 70] + 1
df$cass70[df$lcx >= 70] <- df$cass70[df$lcx >= 70] + 1
df$cass70[df$lm >= 70] <- 2
df$cass70[df$rca >= 70] <- df$cass70[df$rca >= 70] + 1
df$cass70[is.na(df$ang1results)] <- NA

svar <- c(
"uniqueid", 
"stenosis", 
"lm", 
"lad", 
"lcx", 
"rca",
"cass50", 
"cass70"
)

tmp <- df[svar]

### Gensini Score
  # Points for amount stenosis
    # 1 pt > 0%
    # 2 pt > 25%
    # 4 pt > 50%
    # 8 pt > 75%
    # 16 pt > 90%
    # 32 pt = 100%
  # Points are multiplied by lesion importance in circulation
    # 5 x LM
    # 2.5 x prox LAD
    # 2.5 x prox LCX
    # 1.5 x mid LAD
    # 1.0 x RCA | distal LAD | posterolateral (PDA) | obtuse marginal
    # 0.5 x all other segments
  # Gensini score is the sum of all segments

# Appropriately named arteries for scoring
df <- 
  cath %>%
  select(., c(uniqueid, ang1sten1:ang1sten22)) %>%
  mutate_all(., ~replace(., is.na(.), 0)) 

# Need to have overal stenoses points
df[-1] %<>%
  mutate_all(., function(x) {
    case_when(
      x == 100 ~ 32,
      x > 90 ~ 16,
      x > 75 ~ 8,
      x > 50 ~ 4,
      x > 25 ~ 2,
      x >= 0 ~ 1
    )}
  )

# Need to multiply the column by their point values
df %<>%
  dplyr::rename(
    lm = ang1sten1,
    plad1 = ang1sten2,
    plad2 = ang1sten3,
    plcx = ang1sten6,
    mlad = ang1sten4,
    dlad = ang1sten5,
    rca1 = ang1sten9,
    rca2 = ang1sten10,
    rca3 = ang1sten11,
    rca4 = ang1sten12,
    pda = ang1sten13,
    om1 = ang1sten18,
    om2 = ang1sten19,
    om3 = ang1sten20
    )

# Find maximum points per groups of arteries
df$plad <- apply(X = df[c("plad1", "plad2")], MARGIN = 1, FUN = max, na.rm = TRUE)
df$plad[df$plad == -Inf] <- 0

df$rca <- apply(X = df[c("rca1", "rca2", "rca3", "rca4")], MARGIN = 1, FUN = max, na.rm = TRUE)
df$rca[df$rca == -Inf] <- 0

df$om <- apply(X = df[c("om1", "om2", "om3")], MARGIN = 1, FUN = max, na.rm = TRUE)
df$om[df$om == -Inf] <- 0

df %<>% select(., -c(plad1, plad2, rca1, rca2, rca3, rca4, om1, om2, om3))

# Multiple the scores!
df %<>% within(., {
  lm %<>% `*`(5)
  plad %<>% `*`(2.5)
  plcx %<>% `*`(2.5)
  mlad %<>% `*`(1.5)
  rca %<>% `*`(1.0)
  dlad %<>% `*`(1.0)
  pda %<>% `*`(1.0)
  om %<>% `*`(1.0)
  ang1sten7 %<>% `*`(0.5)
  ang1sten8 %<>% `*`(0.5)
  ang1sten14 %<>% `*`(0.5)
  ang1sten15 %<>% `*`(0.5)
  ang1sten16 %<>% `*`(0.5)
  ang1sten17 %<>% `*`(0.5)
  ang1sten21 %<>% `*`(0.5)
  ang1sten22 %<>% `*`(0.5)
})

df$gensini <-
  df[-1] %>%
  rowSums(.)

df$gensini[is.na(cath$ang1results)] <- NA

# Final data set
angio_scores <- inner_join(df[c("uniqueid", "gensini")], tmp, by = "uniqueid")

```

## Mortality and outcomes

Mortality and outcomes was also extracted, although they have not yet been adjudicated or organized by type of outcome event (e.g. MACE versus other).

```{r, include = FALSE}
# Data of mortality
df <- 
  subset(df_num, redcap_event_name %in% c("1year_followup_arm_1", "2yearly_follow_up_arm_1", "5year_followup_arm_1", "10year_follow_up_arm_1"))

# Select vars
df <- df[c("uniqueid", "redcap_event_name", "pstatusyr")]

# Overall patient status
df %>%
  group_by(uniqueid) %>%
  summarise(Status = sum(pstatusyr, na.rm = TRUE))
df <- aggregate(df$pstatusyr, by = list(uniqueid = df$uniqueid), FUN=sum, na.rm = TRUE, na.action = NULL)

# Rename the column
names(df)[2] <- "status"
df$status[df$status > 0] <- 1

# Save
df_outcomes <- df
```


# Depression and Coronary Artery Disease

## Descriptive analysis

The data is described below. We compared all patients by depression score and age first. We suspect there is an interaction between age, gender, and depression, and this is compared/visualized below.

```{r}
# Data
df <- inner_join(df_demo, df_sad, by = "uniqueid")

compareGroups(gend ~ age + phq + sad, data = df) %>%
  createTable(.) %>%
  export2md()
```

The data also has catherization data. This is the breakdown by angiogram score for CAD severity. The table shows the major epicardial vessels with the mean percent stenosis found, along with the major angiographic scoring systems.


```{r}
df <- angio_scores

compareGroups(~ lm + lad + lcx + rca + cass50 + cass70 + gensini, data = df) %>%
  createTable() %>% 
  export2md()
```

All the groups were compared by gender, age, sex, depression, and presence of stenoses on catherization.

```{r}
df <- inner_join(df_demo, df_sad, by = "uniqueid") %>%
  inner_join(., angio_scores, by = "uniqueid")

# Comparison by age
compareGroups(ageBin ~ gend + phq + cass50 + cass70 + gensini, data = df) %>%
  createTable() %>% 
  export2md(., caption = "Characteristics by Age Group")

# Comparison by sex
compareGroups(gend ~ age + phq + cass50 + cass70 + gensini, data = df) %>%
  createTable() %>% 
  export2md(., caption = "Characteristics by Sex")

# Comparison by sad
compareGroups(sad ~ age + gend + cass50 + cass70 + gensini, data = df) %>%
  createTable() %>% 
  export2md(., caption = "Characteristics by Depression status")

# Comparison by reported obstructive disease
compareGroups(stenosis ~ age + gend + phq + cass50 + cass70 + gensini, data = df) %>%
  createTable() %>% 
  export2md(., caption = "Characteristics by obstructive CAD")
```

## Relationship of gender and depression with CAD

Amit's JAHA paper suggested that young women with depression were more susceptible to CAD. 

_"After multivariable adjustment for CAD risk factors, depressive symptoms predicted CAD presence in women aged ≤55 years (odds ratio=1.07 95% confidence interval [CI] 1.02 to 1.13 per 1 point increase in PHQ-9 score), but not in men aged ≤55 years or women aged >55 years."_ ~ Shah et al 2014

TO re-establish this relationship with roughly ~7000 patients, double that of the group in 2014, we first have to check the association of depression with CAD. Per this plot, no obvious correlation by comparing two variables exists.

```{r correlation map}
df <- inner_join(df_sad, angio_scores, by = "uniqueid") %>%
  inner_join(., df_demo, by = "uniqueid")

# Correlation
ggcorr(df[c("age", "gend", "phq", "sad", "stenosis", "gensini", "cass50", "cass70")], palette = "RdBu", name = "rho", label = TRUE)
```

However, the relationship of depression and CAD may be modified by age and sex. We can explore sex and age effects on depression first.

```{r depression age and gender}
df <- inner_join(df_sad, angio_scores, by = "uniqueid") %>%
  inner_join(., df_demo, by = "uniqueid")

# Histogram
ggplot(na.omit(df), aes(x = gend:ageBin, fill = sad)) +
  geom_bar(position = "fill") +
  scale_y_continuous(labels = scales::percent) +
  scale_fill_viridis_d(option = "cividis")
```

Although it seems like younger women have more depression, its not a huge difference like in the 2014 data set. Lets look at the visual differences as it depression predicts CAD.

```{r plot by age and sex}
df <- inner_join(df_sad, angio_scores, by = "uniqueid") %>%
  inner_join(., df_demo, by = "uniqueid")

# Correlation / patterns
ggplot(na.omit(df), aes(x = phq, y = gensini, color = gend)) +
  facet_wrap(~ageBin) + 
  geom_smooth(method = "lm")
```

This pattern shows the largest difference in slopes with younger women compared to younger men. The pattern in old age does not show an interaction at all. There may be some pattern with middle age, but its hard to tell. With young age, men versus women, likely shows an effect modification. So, we should do the regression analyses here.

```{r, results = "asis"}
df <- inner_join(df_sad, angio_scores, by = "uniqueid") %>%
  inner_join(., df_demo, by = "uniqueid")

# Regression analyses
m <- lm(gensini ~ phq, data = df)
n <- lm(gensini ~ phq + gend, data = df)
o <- lm(gensini ~ phq + gend + ageBin, data = df)
p <- lm(gensini ~ phq + gend*ageBin, data = df)
stargazer(m, n, o, p, header = FALSE,
          font.size = "tiny")
```


