---
title: Biobank Team Project
subtitle: Cognitive Impairment is Increased by Comorbid Diabetes and Coronary Artery Disease
author: Biobank Team
date: January 21, 2020
output: html_document
---

# Setup

Basic set up for the upcoming project. This first chunk is for libraries.

```{r, basic set up}
# Sources the file, in the same folder, to ru the command in that file
library(tidyverse)
library(rmarkdown)
library(knitr)
library(magrittr)
```

# Data Intake

This second chunk is to read in the data.

```{r, data intake}
# Object "df_raw" is on the left
# read_csv is from the tidyverse packages, it will read in comma separated files
df_raw <- read_csv("cog.csv", col_names = TRUE, progress = show_progress())

head(df_raw)
```

# Data Tidying

## Covariates

Here, we will identify which covariates we want to use for our modeling. 

```{r}
# What covariates do we need?

# Age
# Race
# Sex
# BMI
# Smoking
# Alcohol
# Exercise
# Sleep
# Educational level
# Chest pain / angina
# Medications (Aspirin, clopidogrel, statins, beta blockers, other cardiac meds)
# Stroke/CVA
# CABG
# H/o myocardial infarction
# H/o PAD
# Diabetes, A1c, random blood sugar
# Hypertension
# HLD
# Depression / PHQ9
```


## MOCA scores
One thing we will have to do is come up with a score breakdown for the MOCA.

- are there subsections of hte MOCA that each should be scored?
- Are there total score groupings or categories?
- Can you score the MOCA if its incomplete?

```{r, show how to add a column}
# OUr data set
df <- df_raw

# First we need to subset/look at the data

# This is to select just a single column
subset(df, select = Total)

# However, how do we find which patients gave NA values
subset(df, select = c(UniqueID, Total))

# Now we can pipe that subset into our command to remove NAs
subset(df, select = c(UniqueID, Total)) %>%
	na.omit()

# We can finally assign it to a variable named tmp
tmp <- subset(df, select = c(UniqueID, Total)) %>%
	na.omit()

# Lets add that new column for score category
# Score categories are >26, 18 to 25 is mild, 10 to 17 is moderate, <10 is severe
tmp$category <- NA
tmp$category[tmp$Total > 25] <- "Normal"
tmp$category[tmp$Total >= 18 & tmp$Total <= 25] <- "Mild"
tmp$category[tmp$Total >= 10 & tmp$Total <= 17] <- "Moderate"
tmp$category[tmp$Total < 10] <- "Severe"

### How do we save/merge this data back in?
nrow(tmp) # This tells us there are only 67 patients
nrow(df) # This tells us original data is 87 patients long

# We can merge this back in using a left join
moca <- full_join(df, tmp[,c("UniqueID", "category")], by = "UniqueID")
```

