---
title: Biobank Team Project
subtitle: Cognitive Impairment is Increased by Comorbid Diabetes and Coronary Artery Disease
author: Biobank Team
date: January 21, 2020
output: html_document
---

# Overview

## Research Question

- __Do coronary disease and peripheral vascular disease have teh same effects on cognitive function?__

## R Libraries 

Basic set up for the upcoming project. This first chunk is for libraries.

```{r, basic set up}
# Sources the file, in the same folder, to ru the command in that file
library(tidyverse)
library(rmarkdown)
library(knitr)
library(magrittr)
```

## Steps needed

1. Get a data set of CAD/MOCA done at the same time.
2. Clean the data set for both CAD and MOCA scores
3. Add in the chronic conditions and baseline variables that were __only__ taken at enrollment 1

TODO:

- create the covariate dataset
- recreate the MOCA dataset

# Data Intake


## Reading In Data

```{r}
# This a direct export from REDCAP
raw <- read_csv("../team_project/moca-cad.csv", col_names = TRUE)

# Lets make a clean enrollment number
raw$enrollment <- recode(raw$redcap_event_name, enrollment_1_arm_1 = 1, enrollment_2_arm_1 = 2, enrollment_3_arm_1 = 3, enrollment_4_arm_1 = 4)

# Processing data set
proc <- raw

# Limit to just enrollments with caths
proc <- subset(proc, enrollment %in% c(1:4))

# Limit it to just enrollments with MOCA scores
proc <- subset(proc, !is.na(mocatotal))

# Key data set of barebones information
keydata <- subset(proc, select = c(uniqueid, enrollment))
```

# Data Cleaning

## Coronary artery disease data

We can read in the cath data from every visit (including repeat visits) and turn it into angiography scoring (CASS and Gensini).

```{r}
# Pull in the source file called "cath-data.R" and use it to get our angiogram scores
# Uses the "raw" data from above
source("../team_project/cath-data.R")

# This generates the data frame angioscores
head(angio_scores)
```

## Cognitive data only

```{r}
# Dataset trimed
cog <- subset(proc, select = c(uniqueid, enrollment, mocatotal))

```


## Peripheral vascular disease data

Dr. Hajjar is the owner of the PAD data set. That data set includes MOCA scores. How do we measure PVD? These are not Biobank patients.

We've emailed through Quyyumi to Hajjar to receive this data set.

## Covariates

```{r}
# Add in the covariates
baseline <- subset(raw, enrollment == 1)

# Add in covariates to make a demographic/covariate dataset
covar <- left_join(keydata, baseline[c("uniqueid", "gend", "race")], by = "uniqueid")

# What covariates do we need?

# Age
# Race
# Sex
# BMI
# Smoking
# Alcohol
# Exercise
# Sleep
# Educational level
# Chest pain / angina
# Medications (Aspirin, clopidogrel, statins, beta blockers, other cardiac meds)
# Stroke/CVA
# CABG
# H/o myocardial infarction
# H/o PAD
# Diabetes, A1c, random blood sugar
# Hypertension
# HLD
# Depression / PHQ9
```

## MOCA scores

One thing we will have to do is come up with a score breakdown for the MOCA.

- are there subsections of hte MOCA that each should be scored?
- Are there total score groupings or categories?
- Can you score the MOCA if its incomplete?

```{r, show how to add a column}
# OUr data set
df <- cog

# First we need to subset/look at the data

# This is to select just a single column
subset(df, select = Total)

# However, how do we find which patients gave NA values
subset(df, select = c(UniqueID, Total))

# Now we can pipe that subset into our command to remove NAs
subset(df, select = c(UniqueID, Total)) %>%
	na.omit()

# We can finally assign it to a variable named tmp
tmp <- subset(df, select = c(UniqueID, Total)) %>%
	na.omit()

# Lets add that new column for score category
# Score categories are >26, 18 to 25 is mild, 10 to 17 is moderate, <10 is severe
tmp$category <- NA
tmp$category[tmp$Total > 25] <- "Normal"
tmp$category[tmp$Total >= 18 & tmp$Total <= 25] <- "Mild"
tmp$category[tmp$Total >= 10 & tmp$Total <= 17] <- "Moderate"
tmp$category[tmp$Total < 10] <- "Severe"

### How do we save/merge this data back in?
nrow(tmp) # This tells us there are only 67 patients
nrow(df) # This tells us original data is 87 patients long

# We can merge this back in using a left join
moca <- full_join(df, tmp[,c("UniqueID", "category")], by = "UniqueID")
```

